---
title: "Machine learning Models"
format: html
---
# **Machine Learning Models 1**

# Load packages, define the seed, and read the data
```{r}
# Core modeling and plotting packages
library(tidymodels)
library(ggplot2)
library(glmnet)
library(ranger)
library(corrplot)
library(here)

# Set the random-number seed requested in the instructions
rngseed <- 1234
set.seed(rngseed)

# Load the attached RDS file
analysis_df <- readRDS(here("ml-models-exercise","analysis_df.rds"))

# Quick check of the data structure
str(analysis_df)
```
# Combine RACE categories 7 and 88 into category 3 and created BMI. 
```{r}
analysis_df <- analysis_df |>
  mutate(
    # Ensure categorical variables are stored as factors
    DOSE = factor(DOSE),
    SEX  = factor(SEX),
    RACE = as.character(RACE),
    RACE = if_else(RACE %in% c("7", "88"), "3", RACE),
    RACE = factor(RACE),

    # Correct BMI formula for weight in kg and height in meters
    BMI = WT / (HT^2)
  ) |>
  filter(!is.na(Y))

# Quick checks
summary(analysis_df$BMI)
str(analysis_df)
```
# Correlation plot for continuous variables
```{r}
continuous_df <- analysis_df |>
  select(where(is.numeric))

cor_mat <- cor(continuous_df, use = "pairwise.complete.obs")

corrplot(
  cor_mat,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  diag = FALSE
)
```

# Setting up recipes and workflows for the three models
```{r}
# Recipe for linear regression and LASSO
lm_lasso_recipe <- recipe(Y ~ ., data = analysis_df) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_lincomb(all_predictors()) |>
  step_normalize(all_numeric_predictors())

# Recipe for random forest
rf_recipe <- recipe(Y ~ ., data = analysis_df) |>
  step_zv(all_predictors())

# Model 1: ordinary linear regression
lm_spec <- linear_reg() |>
  set_engine("lm")

# Model 2: LASSO with fixed penalty = 0.1 for now
lasso_spec <- linear_reg(
  penalty = 0.1,
  mixture = 1
) |>
  set_engine("glmnet")

# Model 3: random forest with ranger
rf_spec <- rand_forest() |>
  set_engine("ranger", seed = rngseed) |>
  set_mode("regression")

# Workflows
lm_wf <- workflow() |>
  add_recipe(lm_lasso_recipe) |>
  add_model(lm_spec)

lasso_wf <- workflow() |>
  add_recipe(lm_lasso_recipe) |>
  add_model(lasso_spec)

rf_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_spec)
```
## Fit the three models
```{r}
lm_fit <- fit(lm_wf, data = analysis_df)
lasso_fit <- fit(lasso_wf, data = analysis_df)
rf_fit <- fit(rf_wf, data = analysis_df)
```

## Make predictions on the full dataset and compute RMSE

```{r}
# Helper function to add predictions and compute RMSE
get_apparent_rmse <- function(fitted_wf, data, model_name) {
  preds <- predict(fitted_wf, new_data = data) |>
    bind_cols(data |> select(Y))

  rmse_val <- rmse(preds, truth = Y, estimate = .pred)
  rmse_val$model <- model_name

  list(preds = preds, rmse = rmse_val)
}

lm_results <- get_apparent_rmse(lm_fit, analysis_df, "Linear model")
lasso_results <- get_apparent_rmse(lasso_fit, analysis_df, "LASSO (penalty = 0.1)")
rf_results <- get_apparent_rmse(rf_fit, analysis_df, "Random forest")

rmse_table <- bind_rows(
  lm_results$rmse,
  lasso_results$rmse,
  rf_results$rmse
) |>
  select(model, .metric, .estimate)

rmse_table
```
# Observed versus predicted plots
```{r}

plot_obs_pred <- function(pred_df, model_title) {
  ggplot(pred_df, aes(x = Y, y = .pred)) +
    geom_point(alpha = 0.6) +
    geom_abline(intercept = 0, slope = 1, linetype = 2) +
    labs(
      title = paste("Observed vs Predicted:", model_title),
      x = "Observed Y",
      y = "Predicted Y"
    ) +
    theme_minimal()
}

plot_obs_pred(lm_results$preds, "Linear model")
plot_obs_pred(lasso_results$preds, "LASSO (penalty = 0.1)")
plot_obs_pred(rf_results$preds, "Random forest")
```
# Tune the LASSO using apparent() resamples

```{r}
# Create a tunable LASSO model
lasso_tune_spec <- linear_reg(
  penalty = tune(),
  mixture = 1
) |>
  set_engine("glmnet")

lasso_tune_wf <- workflow() |>
  add_recipe(lm_lasso_recipe) |>
  add_model(lasso_tune_spec)

# Grid of 50 penalty values on the log10 scale
lasso_grid <- tibble(
  penalty = 10^seq(log10(1e-5), log10(1e2), length.out = 50)
)

# Apparent resample object
apparent_resamples <- apparent(analysis_df)

# Tune over the grid
lasso_apparent_tuned <- tune_grid(
  lasso_tune_wf,
  resamples = apparent_resamples,
  grid = lasso_grid,
  metrics = metric_set(rmse),
  control = control_grid(save_pred = TRUE)
)

# For apparent() there is only one resample, so keep the unsummarized results
lasso_metrics_apparent <- collect_metrics(lasso_apparent_tuned, summarize = FALSE) |>
  arrange(.estimate)

lasso_metrics_apparent
```

# Manual Plot of LASSO tuning results
```{r}
# autoplot was not working for me so I made my own plot from the lasso apparent resampling Matrics 
ggplot(lasso_metrics_apparent, aes(x = penalty, y = .estimate)) +
  geom_line() +
  geom_point() +
  scale_x_log10() +
  labs(
    title = "LASSO tuning results using apparent resampling",
    x = "Penalty (log scale)",
    y = "RMSE"
  ) +
  theme_minimal()

```


## Tune the random forest using apparent() resamples

```{r}
rf_tune_spec <- rand_forest(
  trees = 300,
  mtry = tune(),
  min_n = tune()
) |>
  set_engine("ranger", seed = rngseed) |>
  set_mode("regression")

rf_tune_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_tune_spec)

rf_grid <- grid_regular(
  mtry(range = c(1L, 7L)),
  min_n(range = c(1L, 21L)),
  levels = 7
)

rf_apparent_tuned <- tune_grid(
  rf_tune_wf,
  resamples = apparent_resamples,
  grid = rf_grid,
  metrics = metric_set(rmse),
  control = control_grid(save_pred = TRUE)
)

rf_metrics_apparent <- collect_metrics(rf_apparent_tuned, summarize = FALSE) |>
  arrange(.estimate)

rf_metrics_apparent
```
# Manual plot of random forest apparent tuning results
```{r}
ggplot(rf_metrics_apparent, aes(x = factor(mtry), y = factor(min_n), fill = .estimate)) +
  geom_tile() +
  labs(
    title = "Random forest tuning results using apparent resampling",
    x = "mtry",
    y = "min_n",
    fill = "RMSE"
  ) +
  theme_minimal()
```



## Proper tuning with repeated cross-validation


```{r}
set.seed(rngseed)

cv_5x5 <- vfold_cv(
  analysis_df,
  v = 5,
  repeats = 5
)

cv_5x5
```

# Re-tune the LASSO with repeated cross-validation

```{r}
lasso_cv_tuned <- tune_grid(
  lasso_tune_wf,
  resamples = cv_5x5,
  grid = lasso_grid,
  metrics = metric_set(rmse)
)

lasso_cv_tuned
```

```{r}
autoplot(lasso_cv_tuned)
```

## Re-tune the random forest with repeated cross-validation

```{r}
rf_cv_tuned <- tune_grid(
  rf_tune_wf,
  resamples = cv_5x5,
  grid = rf_grid,
  metrics = metric_set(rmse)
)

rf_cv_tuned
```

```{r}
autoplot(rf_cv_tuned)
```

# show the best settings found by cross-validation

```{r}
show_best(lasso_cv_tuned, metric = "rmse")
```

```{r}
show_best(rf_cv_tuned, metric = "rmse")
```
