---
title: "Fitting Basic Statistical Models"
format: html
---
# Data processing and exploration
```{r}
# Load libraries
library(readr)
library(ggplot2)
library(dplyr)
library(here)
```

```{r}
# Load the CSV file
Mavoglurant_data <- read_csv(here("fitting-exercise", "Mavoglurant_A2121_nmpk.csv"))

# View first rows
head(data)

# Check column types
glimpse(data)
```


```{r}
# Faceted Plot with Different Color per Dose
ggplot(Mavoglurant_data, aes(x = TIME, y = DV, group = ID, color = DOSE)) +
  geom_line(alpha = 0.6) +
  facet_wrap(~ DOSE) +
  labs(
    title = "Outcome (DV) Over Time Stratified by Dose",
    x = "Time",
    y = "DV",
    color = "Dose"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
# Keep only observations where OCC == 1
Mavoglurant_data_occ1 <- Mavoglurant_data %>%
  filter(OCC == 1)
```

```{r}
# Check that only OCC = 1 remains
unique(Mavoglurant_data_occ1$OCC)

# Compare number of rows before and after filtering
nrow(Mavoglurant_data)
nrow(Mavoglurant_data_occ1)
```

```{r}
# Remove observations where TIME == 0
data_no_time0 <- Mavoglurant_data_occ1 %>%
  filter(TIME != 0)

# Compute sum of DV for each individual
Y_df <- data_no_time0 %>%
  group_by(ID) %>%
  summarize(Y = sum(DV, na.rm = TRUE)) %>%
  ungroup()
```

```{r}
# Create Dataset with TIME = 0 Only
time0_df <- Mavoglurant_data_occ1 %>%
  filter(TIME == 0)
```

```{r}
# Join the Two Data Frames
final_df <- left_join(time0_df, Y_df, by = "ID")
```

```{r}
# Convert RACE and SEX to factors and keep only selected variables
analysis_df <- final_df %>%
  mutate(
    RACE = as.factor(RACE),
    SEX  = as.factor(SEX)
  ) %>%
  select(Y, DOSE, AGE, SEX, RACE, WT, HT)
```

```{r}
# Check structure
str(analysis_df)

# Check dimensions
dim(analysis_df)
```

# EDA


```{r}
# Load required packages
library(tidyverse)
library(GGally)

# Overall summary statistics
analysis_df %>%
  summarize(
    n = n(),
    mean_Y = mean(Y, na.rm = TRUE),
    sd_Y   = sd(Y, na.rm = TRUE),
    mean_AGE = mean(AGE, na.rm = TRUE),
    mean_WT  = mean(WT, na.rm = TRUE),
    mean_HT  = mean(HT, na.rm = TRUE)
  )
```

```{r}
# Summary by DOSE
analysis_df %>%
  group_by(DOSE) %>%
  summarize(
    n = n(),
    mean_Y = mean(Y, na.rm = TRUE),
    sd_Y   = sd(Y, na.rm = TRUE)
  )
# Counts for Categorical Variables
table(analysis_df$SEX)
table(analysis_df$RACE)
table(analysis_df$DOSE)
```

```{r}
# Scatterplots and Boxplots
# Y vs Continuous Predictors
# Y vs AGE
ggplot(analysis_df, aes(x = AGE, y = Y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Total Drug (Y) vs Age")

# Y vs WT
ggplot(analysis_df, aes(x = WT, y = Y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Total Drug (Y) vs Weight")

# Y vs HT
ggplot(analysis_df, aes(x = HT, y = Y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Total Drug (Y) vs Height")
```
```{r}
# Y vs Categorical Predictors
# Y by DOSE
ggplot(analysis_df, aes(x = DOSE, y = Y)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Total Drug (Y) by Dose")

# Y by SEX
ggplot(analysis_df, aes(x = SEX, y = Y)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Total Drug (Y) by Sex")

# Y by RACE
ggplot(analysis_df, aes(x = RACE, y = Y)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Total Drug (Y) by Race")
```

```{r}
# Correlation Matrix (Continuous Variables Only)
# Select continuous variables
continuous_vars <- analysis_df %>%
  select(Y, AGE, WT, HT)

# Correlation matrix
cor(continuous_vars, use = "complete.obs")
GGally::ggpairs(
  continuous_vars,
  title = "Pairwise Relationships Between Continuous Variables"
)
```

# Model Fitting 

```{r}
# load libraries 
library(tidyverse)
library(tidymodels)
```

```{r}
# Ensure categorical variables are coded as factors
analysis_df <- analysis_df %>%
  mutate(
    DOSE = as.factor(DOSE),
    SEX  = as.factor(SEX),
    RACE = as.factor(RACE)
  )

set.seed(123)    # Ensures you get the same split every time you run
```
```{r}
# Create 10-fold cross-validation object
cv_folds <- vfold_cv(
  analysis_df,      # dataset
  v = 10,           # number of folds
  strata = DOSE     # keep dose distribution balanced across folds
)
cv_folds_sex <- vfold_cv(       # Create 10-fold cross-validation splits for the SEX models
  analysis_df,                  # Data to split into folds
  v = 10,                       # Number of folds
  strata = SEX                  # Keep SEX class balance similar across folds
)
```

# CONTINUOUS OUTCOME (Y)- LINEAR REGRESSION

```{r}
# Linear regression models (Outcome = Y)

# Model specification
lm_spec <- linear_reg() %>% # Define a linear regression model
  set_engine("lm")          # Use base R's lm() function to fit the model
```
```{r}
# Model 1: Predict Y using DOSE
rec_y_dose <- recipe(               # Create preprocessing recipe
  Y ~ DOSE,                         # Define formula: Y predicted by DOSE
  data = analysis_df                # Use the full dataset to define variables
) %>%
  step_dummy(all_nominal_predictors()) # Convert categorical predictors (DOSE) into dummy variables

wf_y_dose <- workflow() %>%         # Create a workflow object
  add_recipe(rec_y_dose) %>%        # Add preprocessing steps from the recipe
  add_model(lm_spec)                # Add the linear regression model

res_y_dose <- fit_resamples(        # Fit model repeatedly across cross-validation folds
  wf_y_dose,                        # Workflow containing model and preprocessing
  resamples = cv_folds,             # Use the 10-fold cross-validation object
  metrics = metric_set(rmse, rsq)   # Compute RMSE and R-squared performance metrics
)

metrics_y_dose <- collect_metrics(res_y_dose) # Collect the averaged performance metrics from all folds

```

```{r}
# Model 2: Predict Y using all predictors
rec_y_all <- recipe(                # Create recipe for full model
  Y ~ DOSE + AGE + SEX + RACE + WT + HT,  # Predict Y using all listed predictors
  data = analysis_df                # Use dataset to determine variable types
) %>%
  step_dummy(all_nominal_predictors()) # Convert categorical variables into dummy indicators

wf_y_all <- workflow() %>%          # Create workflow for full model
  add_recipe(rec_y_all) %>%         # Add preprocessing recipe
  add_model(lm_spec)                # Add linear regression model specification

res_y_all <- fit_resamples(         # Perform cross-validated model fitting
  wf_y_all,                         # Workflow containing recipe and model
  resamples = cv_folds,             # Use the previously defined 10 folds
  metrics = metric_set(rmse, rsq)   # Evaluate using RMSE and R-squared
)

metrics_y_all <- collect_metrics(res_y_all) # Extract averaged cross-validation results
```
```{r}
# Printing results for linear regression models
cat("\n Linear Regression Results (Cross-Validation) \n") 

cat("\nModel 1: Y ~ DOSE\n")   
print(metrics_y_dose)          

cat("\nModel 2: Y ~ DOSE + AGE + SEX + RACE + WT + HT\n") 
print(metrics_y_all)           
```

# Logistic regression models (Outcome = SEX)
```{r}
# Define logistic regression model specification
log_spec <- logistic_reg() %>%  # Specify logistic regression model
  set_engine("glm")             # Use base R's glm() function with binomial link
```
```{r}
# Model 3: Predict SEX using DOSE
rec_s_dose <- recipe(           # Create preprocessing recipe
  SEX ~ DOSE,                   # Predict SEX using DOSE
  data = analysis_df            # Use dataset to determine variable roles
) %>%
  step_dummy(all_nominal_predictors()) # Convert DOSE into dummy variables

wf_s_dose <- workflow() %>%     # Build workflow for logistic regression
  add_recipe(rec_s_dose) %>%    # Add preprocessing recipe
  add_model(log_spec)           # Add logistic regression model

res_s_dose <- fit_resamples(    # Perform cross-validation fitting
  wf_s_dose,                    # Workflow object
  resamples = cv_folds,         # Use same 10-fold cross-validation
  metrics = metric_set(accuracy, roc_auc) # Evaluate using accuracy and ROC-AUC
)

metrics_s_dose <- collect_metrics(res_s_dose) # Extract averaged performance metrics
```
```{r}
# Model 4: Predict SEX using all predictors
rec_s_all <- recipe(            # Create preprocessing recipe
  SEX ~ DOSE + AGE + RACE + WT + HT + Y, # Predict SEX using these predictors
  data = analysis_df            # Use dataset to determine variable types
) %>%
  step_dummy(all_nominal_predictors()) # Convert categorical predictors to dummy variables

wf_s_all <- workflow() %>%      # Create workflow object
  add_recipe(rec_s_all) %>%     # Add preprocessing recipe
  add_model(log_spec)           # Add logistic regression model

res_s_all <- fit_resamples(     # Fit logistic model across cross-validation folds
  wf_s_all,                     # Workflow containing model and preprocessing
  resamples = cv_folds,         # Use the same folds
  metrics = metric_set(accuracy, roc_auc) # Evaluate classification performance
)

metrics_s_all <- collect_metrics(res_s_all) # Collect averaged accuracy and ROC-AUC values

# Print results for logistic regression models
cat("\n Logistic Regression Results (Cross-Validation) \n") 

cat("\nModel 3: SEX ~ DOSE\n")   
print(metrics_s_dose)           

cat("\nModel 4: SEX ~ DOSE + AGE + RACE + WT + HT + Y\n") # Identify the fourth model
print(metrics_s_all)             
```


# This code uses the tidymodels framework with 10 fold cross validation to evaluate the models. Instead of splitting the dataset into one training set and one testing set, the data are divided into ten smaller groups (folds). The model is trained on nine of these groups and tested on the remaining one, and this process is repeated ten times so that each group is used as the test set once. The performance metrics (RMSE, R², accuracy, and ROC-AUC) are then averaged across all ten runs.


# Interpretation of Model Results

# For the linear regression models predicting total drug exposure (Y), the model using only DOSE had an RMSE of about 667 and an R² of 0.574. This means that dose alone explains about 57% of the variability in Y, which suggests that DOSE is an important predictor of total drug exposure. When the other predictors (AGE, SEX, RACE, WT, and HT) were added to the model, the RMSE decreased slightly to around 626 and the R² increased to 0.601. This shows a small improvement in model performance, meaning the additional variables help explain a little more of the variation in Y, but DOSE still appears to be the most important factor.

# For the logistic regression models predicting SEX, the model using only DOSE had an accuracy of about 0.87 but a low ROC-AUC of 0.567. This suggests that while the model classified many observations correctly, DOSE alone does not do a great job distinguishing between the two SEX groups. When all predictors were included, the model improved quite a bit, with accuracy increasing to about 0.92 and ROC-AUC increasing to 0.836. This indicates that variables like AGE, weight, height, and Y provide useful information for predicting SEX. Overall, the results make sense because dose would naturally affect total drug exposure, while characteristics like body size are more likely to be related to sex differences.



# **Model Improvement Exercise**


```{r}

# Load required packages
library(tidyverse)
library(tidymodels)
# define seed
rngseed <- 1234 
```
```{r}
# Remove the RACE variable

analysis_df_no_race <- analysis_df %>%
  select(-RACE)

# Use glimpse to inspect the dataset
glimpse(analysis_df_no_race)
```
```{r}

# Set seed and split into 75% train / 25% test

set.seed(rngseed)

data_split <- initial_split(analysis_df_no_race, prop = 0.75)

train_data <- training(data_split)
test_data  <- testing(data_split)

# Inspect the split data
glimpse(train_data)
glimpse(test_data)
```
```{r}
# Set seed and split into 75% train / 25% test

set.seed(rngseed)

data_split <- initial_split(analysis_df_no_race, prop = 0.75)

train_data <- training(data_split)
test_data  <- testing(data_split)

# Inspect the split data
glimpse(train_data)
glimpse(test_data)
```
```{r}
# Model 1: Y ~ DOSE

model1_recipe <- recipe(Y ~ DOSE, data = train_data)

model1_workflow <- workflow() %>%
  add_recipe(model1_recipe) %>%
  add_model(lm_spec)

model1_fit <- fit(model1_workflow, data = train_data)
```
```{r}
# Model 2: Y ~ all predictors

model2_recipe <- recipe(Y ~ ., data = train_data)

model2_workflow <- workflow() %>%
  add_recipe(model2_recipe) %>%
  add_model(lm_spec)

model2_fit <- fit(model2_workflow, data = train_data)
```

```{r}
# Compute predictions on the TRAINING data

# Model 1 training predictions
model1_train_preds <- train_data %>%
  bind_cols(predict(model1_fit, new_data = train_data)) %>%
  rename(predicted = .pred) %>%
  select(Y, predicted)

# Model 2 training predictions
model2_train_preds <- train_data %>%
  bind_cols(predict(model2_fit, new_data = train_data)) %>%
  rename(predicted = .pred) %>%
  select(Y, predicted)
```

```{r}
# Compute RMSE on the TRAINING data

model1_train_rmse <- model1_train_preds %>%
  rmse(truth = Y, estimate = predicted) %>%
  mutate(model = "Y ~ DOSE")

model2_train_rmse <- model2_train_preds %>%
  rmse(truth = Y, estimate = predicted) %>%
  mutate(model = "Y ~ all predictors")

# Compare training RMSE
train_rmse_results <- bind_rows(model1_train_rmse, model2_train_rmse) %>%
  select(model, .metric, .estimate)

train_rmse_results
```


```{r}
# Identify best-fitting model on training data

best_model_name <- train_rmse_results %>%
  arrange(.estimate) %>%
  slice(1) %>%
  pull(model)

best_model_name
```

```{r}
# Compute RMSE for the null model on TRAINING data
#    (predicts the mean of Y for all observations)

y_mean <- mean(train_data$Y, na.rm = TRUE)

null_train_preds <- train_data %>%
  mutate(predicted = y_mean) %>%
  select(Y, predicted)

null_train_rmse <- null_train_preds %>%
  rmse(truth = Y, estimate = predicted) %>%
  mutate(model = "Null model") %>%
  select(model, .metric, .estimate)

null_train_rmse
```

```{r}

# Compare RMSE for both fitted models and null model

all_train_rmse <- bind_rows(
  model1_train_rmse %>% select(model, .metric, .estimate),
  model2_train_rmse %>% select(model, .metric, .estimate),
  null_train_rmse
)

all_train_rmse

```
# Discussion on RMSE value 
# The Null Model (baseline) is doing the worst with a huge error of 948. When we moved to Model 1, the error dropped significantly to 702. But Model 2 improved things even further, knocking the error down to 627. Which means using all predictors provided a better result than just using Dose. 
```{r}

# Re-set seed and perform 10-fold cross-validation
set.seed(rngseed)

cv_folds <- vfold_cv(train_data, v = 10)
```

```{r}

# Compute cross-validated RMSE for Model 1

model1_cv_results <- fit_resamples(
  model1_workflow,
  resamples = cv_folds,
  metrics = metric_set(rmse)
)

collect_metrics(model1_cv_results)

```

```{r}
# Compute cross-validated RMSE for Model 2

model2_cv_results <- fit_resamples(
  model2_workflow,
  resamples = cv_folds,
  metrics = metric_set(rmse)
)

collect_metrics(model2_cv_results)
```

```{r}
# Compare cross-validated RMSE for both models

cv_rmse_results <- bind_rows(
  collect_metrics(model1_cv_results) %>% mutate(model = "Y ~ DOSE"),
  collect_metrics(model2_cv_results) %>% mutate(model = "Y ~ all predictors")
) %>%
  select(model, .metric, mean, std_err)

cv_rmse_results
```
# Discussion on RMSE value after CV
# Overall, Model 2 still outperforms Model 1 and the null model after cross-validation, but the gap between the models narrows. This suggests that while Model 2 remains the best predictive model, its advantage is somewhat reduced when evaluated on unseen data.

# This section added by Joe Dainis

## Make dataframes for all of the observed data and plot it
```{r}
#Dataframe of predictions for Model 1
model1fit_df <- train_data %>%
  mutate(predicted = predict(model1_fit, new_data = train_data)$.pred,
         model = "Model 1: DOSE")

#Dataframe of predictions for Model 2
model2fit_df <- train_data %>%
  mutate(predicted = predict(model2_fit, new_data = train_data)$.pred,
         model = "Model 2: All predictors")

#Dataframe of null model predictions
y_mean

null_df <- train_data %>%
  mutate(predicted = y_mean,
         model = "Null model")

#Combine the three dataframes into one
all_predictions <- bind_rows(model1fit_df, model2fit_df, null_df)

#Plot the observed versus predicted Y values from the 3 models
all_predictions_plot <- ggplot(all_predictions, aes(x = Y, y = predicted, color = model)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 5000), ylim = c(0, 5000)) +
  labs(
    title = "Observed vs Predicted Values Across Dose, All Variables and Null Models",
    x = "Observed Y",
    y = "Predicted Y"
  ) +
  theme_minimal()

all_predictions_plot
```

## Plot Residuals for Model 2
```{r}
#Create residuals from predicted - observed for Model 2
model2_residuals <- train_data %>%
  mutate(
    predicted = predict(model2_fit, new_data = train_data)$.pred,
    residuals = predicted - Y
  )

#Plot predicted Y values versus residuals for Model 2
residuals_plot <- ggplot(model2_residuals, aes(x = predicted, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "solid") +
  labs(
    title = "Predicted Values versus Residuals for Model 2",
    x = "Predicted Y",
    y = "Residuals"
  ) +
  ylim(-3000, 3000) +   # symmetric range
  theme_minimal()

residuals_plot
```

# Bootstrap Sampling
```{r}
#Load necessary libraries
library(purrr)

#Set the seed as done before using rngseed
set.seed(rngseed)

#Create 100 samples using the bootstraps function
boot_samples <- bootstraps(train_data, times = 100)

#Fit the model to the bootstrap samples using the map function
boot_predictions <- map(boot_samples$splits, function(split) {
  boot_data <- analysis(split)
  #Fit All Variables Model on bootstrap samples
  fit <- workflow() %>%
    add_recipe(recipe(Y ~ ., data = boot_data)) %>%
    add_model(lm_spec) %>%
    fit(data = boot_data)
  #Make predictions on original training data
  predict(fit, new_data = train_data)$.pred
})

# Convert the generated list into an array for median and 89% CIs
pred_bs <- do.call(rbind, boot_predictions)

#Compute median and 89% Confidence Intervals
preds <- pred_bs |>
  apply(2, quantile, c(0.055, 0.5, 0.945)) |>
  t()

preds_df <- as.data.frame(preds)
colnames(preds_df) <- c("lower", "median", "upper")

#Combine observed and original predictions into single dataframe
final_boot_df <- train_data %>%
  mutate(
    point_estimate = predict(model2_fit, new_data = train_data)$.pred
  ) %>%
  bind_cols(preds_df)

#Plot the observed values and bootstap sampling output
observed_bootstrap_plot <- ggplot(final_boot_df, aes(x = Y)) +
  #Mean of original predictions
  geom_point(aes(y = point_estimate), color = "black", alpha = 0.6) +
  #Median of the bootstrap samples
  geom_point(aes(y = median), color = "blue", alpha = 0.6) +
  #Confidence interval upper and lower bouds of bootstrap samples
  geom_errorbar(aes(ymin = lower, ymax = upper), color = "purple", alpha = 0.3) +
  #Add solid line from intercept
  geom_abline(slope = 1, intercept = 0, linetype = "solid") +
  #Ensure x and y axes are same limits
  coord_cartesian(xlim = c(0, 5000), ylim = c(0, 5000)) +
  labs(
    title = "Correlation Between Observed and Predicted Values",
    x = "Original Observed Y Values",
    y = "Bootstrapped Y Predictions from Fitted Model 2"
  ) +
  theme_minimal()

observed_bootstrap_plot
```

## Interpretation
I see that there is a positive correlation between the observed Y values and the bootstrapped sample Y predictions fitted from our All Variables Model. This positive correlation indicates that the model is properly predicting the observed Y values, as we would see no trend/scattering of points if the model was not correlating with the observed Y values. As well, the Confidence Intervals (CIs) from the bootstrapping reveals that most points have minimal uncertainity.

# Final evaluation using test data

# Test-Set Predictions for Model 2

```{r}
# Use the fitted Model 2 to make predictions on the test data
model2_test_preds <- test_data %>%
  mutate(predicted = predict(model2_fit, new_data = test_data)$.pred)

# compute test RMSE
model2_test_rmse <- model2_test_preds %>%
  rmse(truth = Y, estimate = predicted)

model2_test_rmse
```

```{r}
# Create a training-data prediction data frame
model2_train_plot_df <- train_data %>%
  mutate(
    predicted = predict(model2_fit, new_data = train_data)$.pred,
    dataset = "Training data"
  ) %>%
  select(Y, predicted, dataset)

# Create a test-data prediction data frame
model2_test_plot_df <- test_data %>%
  mutate(
    predicted = predict(model2_fit, new_data = test_data)$.pred,
    dataset = "Test data"
  ) %>%
  select(Y, predicted, dataset)

# Combine training and test predictions into one data frame
model2_train_test_preds <- bind_rows(
  model2_train_plot_df,
  model2_test_plot_df
)

glimpse(model2_train_test_preds)
```

```{r}
# Plot predicted versus observed for both training and test data
model2_train_test_plot <- ggplot(
  model2_train_test_preds,
  aes(x = Y, y = predicted, color = dataset, shape = dataset)
) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 5000), ylim = c(0, 5000)) +
  labs(
    title = "Observed vs Predicted Values for Model 2",
    x = "Observed Y",
    y = "Predicted Y",
    color = "Data set",
    shape = "Data set"
  ) +
  theme_minimal()

model2_train_test_plot
```
# Discussion: 
Model 2 performed best overall, showing the lowest prediction error and closest alignment to the 45-degree line, indicating strong predictive accuracy. The training data points lie slightly closer to the line than the test data, suggesting mild overfitting but still good generalization. The test set results remain consistent with training performance, indicating that Model 2 generalizes reasonably well to new, unseen data.