---
---
############################################
# Data exercise 
# Option 2: Synthetic Data Generation and Exploration
# Author: Austin Hill
# Purpose: 
# 1. Generate a synthetic dataset of individuals
# 2. Build in a known association (height and NBA probability)
# 3. Explore the data visually and with tables
# 4. Fit simple models to recover the association
############################################

```{r}
# Load required libraries
library(tidyverse)
```

############################################
# 1. Generate Synthetic Data
############################################

```{r}
# Set a seed so results are reproducible
# This ensures the same random data is created every time
set.seed(123)
```

```{r}
# Number of individuals
n <- 100
```

```{r}
# Create a synthetic dataset
synthetic_data <- tibble(
  
  # Height in feet:
  # heights range from 5 to 7 feet
  height_ft = runif(n, min = 5, max = 7),
  
  # NBA probability (0 to 1 initially):
  # Taller people are made in this data to have a higher chance
  # creating a linear relationship plus some random noise
  nba_prob = 0.05 + 0.4 * (height_ft - 5) +   
 # height effect
rnorm(n, mean = 0, sd = 0.05)
)
# Make sure probabilities stay between 0 and 1
synthetic_data <- synthetic_data %>%
  mutate(
    nba_prob = pmin(pmax(nba_prob, 0), 1),
  # Convert probability to percentage (0–100)
    nba_percent = nba_prob * 100
  )
```

```{r}
# View the first few rows
head(synthetic_data)
```

############################################
# 2. Explore the Data with Tables
############################################

```{r}
# Summary statistics for height and NBA probability
summary(synthetic_data)

```

```{r}
# Create a simple table showing average NBA chance by height group
synthetic_data %>%
  mutate(
    height_group = case_when(
      height_ft < 5.75 ~ "Shorter",
      height_ft < 6.5  ~ "Medium",
      TRUE             ~ "Taller"
       )
  ) %>%
  group_by(height_group) %>%
  summarise(
    avg_height = mean(height_ft),
    avg_nba_percent = mean(nba_percent)
  )
```

############################################
# 3. Visualize the Data with ggplot2
############################################

```{r}
# Scatter plot of height vs NBA probability
ggplot(synthetic_data, aes(x = height_ft, y = nba_percent)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Height vs Chance of Making the NBA",
    x = "Height (feet)",
    y = "Chance of Making the NBA (%)"
  )
```

```{r}
# Histogram of height
ggplot(synthetic_data, aes(x = height_ft)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Distribution of Height",
    x = "Height (feet)",
    y = "Person Count"
  )

# Histogram of NBA probability
ggplot(synthetic_data, aes(x = nba_percent)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Distribution of NBA Probability",
    x = "Chance of Making the NBA (%)",
    y = "Person Count"
  )
```
############################################
# 4. Fit Simple Models
############################################

```{r}
# Linear model:
# Predict NBA percentage using height
lm_model <- lm(nba_percent ~ height_ft, data = synthetic_data)
# View model results
summary(lm_model)
```

# Interpretation:
# The coefficient for height_ft is positive,
# this makes sense because I built the data so taller people have a higher chance.

############################################
# 5. Logistic Model 
############################################

```{r}
# Creating a binary outcome:
# Did the person make the NBA? (Yes/No)
# made it a 50/50 shot randomly generated using the prob. data 
synthetic_data <- synthetic_data %>%
  mutate(
    made_nba = rbinom(n, size = 1, prob = nba_prob)
  )

# Fit a logistic regression model
glm_model <- glm(made_nba ~ height_ft,data = synthetic_data,family = binomial)

# View model results
summary(glm_model)
# The Big Takeaway: Height Matters (A Lot) in making the NBA. 
```
############################################
# 6. Compare Model Fits
############################################

```{r}
# Plot predicted probabilities from the logistic model
synthetic_data %>%
  mutate(
    predicted_prob = predict(glm_model, type = "response")
  ) %>%
  ggplot(aes(x = height_ft, y = predicted_prob)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Predicted Probability of Making the NBA by Height",
    x = "Height (feet)",
    y = "Predicted Probability"
  )
```
############################################
# 6. Compare Model Fits
############################################

```{r}
# Plot predicted probabilities from the logistic model
synthetic_data %>%
  mutate(
    predicted_prob = predict(glm_model, type = "response")
  ) %>%
  ggplot(aes(x = height_ft, y = predicted_prob)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Predicted Probability of Making the NBA by Height",
    x = "Height (feet)",
    y = "Predicted Probability"
  )

```
# Overall Notes:
# - The linear model captures the overall trend
# - The logistic model is more appropriate for a yes/no outcome
# - Both models successfully recover the association that taller individuals have a higher chance of making the NBA

# Disclaimer: AI was used to assist with this excercise 